Compiling Python Modules to Native Parallel Modules Using Pythran and OpenMP Annotations

### 2013-10-27 doc/papers/sc2013/paper.tex
SC2013: update taking reviews into account
Still incomplete, missing points are added as comments
(look for FIXME)

Since its birth in December 1989, the Python language~\cite{rossum97} has proved
to be useful in various domains, ranging from system administration to web
services, thanks to its dynamicity, expressiveness, rich ecosystem and
``batteries included'' standard library. It is also widely used in scientific
computing~\cite{Oliphant2007}, thanks to the SciPy~\cite{scipy} project
which provides a set of scientific and numeric tools, ranging from linear
algebra to ordinary differential equation solvers or generic algorithms.

@@

To allow the user to write native functions without having to write the
boilerplate glue code that turns Python object into native structures back and
forth, the SciPy package provides the \texttt{weave} module. It makes it
possible to bundle C code snippets into Python code. They are then compiled and
loaded at runtime. This hybrid approach, based on a mix of interpreted and
native code, is getting widespread in the Python landscape. It is also the
recommended approach to write applications that make use of fine grain
parallelism, as Python parallelism support is only suitable for coarse-grain
parallelism.

@@

\begin{lstlisting}[float, label={lst:buffon-python}, caption={Implementation of
sequential and parallel version of the Buffon algorithm in Python.}]
def buffon(darts):
  hits = 0
  for i in xrange (0, darts):
    x, y = random(), random()
    dist = sqrt(pow(x, 2) + pow(y, 2))
    if dist <= 1.0:
      hits += 1.0
  # hits / throws = 1/4 Pi
  pi = 4 * (hits / darts)
  return pi

from threading import Thread
from multiprocessing import Queue
def threaded_buffon(d):
  def work(darts, queue):
    queue.put(pi_estimate(darts))
  n = 4
  q = Queue()
  threads = [
    Thread(target=work, args=(d//n, q)),
    Thread(target=work, args=(d//n, q)),
    Thread(target=work, args=(d//n, q)),
    Thread(target=work, args=(d//n, q)),
  ]
  map(Thread.start, threads)
  map(Thread.join, threads)
  return sum(q.get() for _ in threads)/n

from multiprocessing import Pool
def multi_buffon(darts):
  n = 4
  p = Pool(n)
  return sum(p.map(buffon, darts//n))/n
\end{lstlisting}
% Mehdi: I don't get the [darts//4]*4 at the last line of the listing

There have also been several approaches to replace the GIL by Transactional
Memories~\cite{Riley2006,Tabba2010} but none of them made their way to the
mainstream interpreters. As a consequence, Python developers need to write
multi-threaded native modules in order to fully benefit from multiple cores.
This leads to a kind of computations referred as \emph{hybrid computations}.

@@

Cython~\cite{cython2010} is an hybrid Python/C dialect. It extends the Python
syntax with typing informations, calls to native functions from third party
libraries, and a limited set of parallelism constructs, such as the possibility
to define parallel loops, but no task parallelism. When possible, it unboxes
Python variables to improve performance. Without type annotations, the
performance improvement is not terrific, but given enough type annotation,
Cython can generates code that runs as fast as its C equivalent, while
maintaining a syntax close to Python.

PLW~\cite{dongara2007} and \texttt{Scipy.weave} both propose an other approach
that directly mixes Python with C, using raw strings to hold the C code. PLW
also supports parallel directives that are limited to parallel for loops.
PyCUDA and PyOpenCL~\cite{klockner2012} also target accelerators by mixing
Python with kernels embedded as raw strings containing accelerator code.

The main drawback of these approaches is that they require to modify in a
backward-incompatible way the original code. They require to learn a new
dialect, and the long-term preservation of this investment is not ensured.
Moreover, there is no fall back if the code were to be deployed in an
environment where the parallelizing tool/module is not available. For instance,
once a code has been ported to PyCUDA and transformed into CUDA code embedded in
Python strings, there is no way back and the developer has to maintain two
versions of the algorithm.

@@

Most backward-compatible approaches also require to modify the input program.
They do not extend the Python language, but restrict it. As a consequence, they
remain compatible with the original language and do not suffer from the
drawbacks of the previous approaches. They also benefit from existing tools
associated to the language.

@@

At the AST level, it means that the Python AST is first reduced to a tree where
all nodes not available in C++ are transformed into a compatible representation.
For instance, list comprehension expressions are turned into function calls or
tuple unpacking is turned in multiple assignments. During this transformation
process, an expression is always transformed into a single expression and a
statement is always transformed into a single statement. Then this AST is
converted into a C++ AST meant to be pretty-printed.

@@

The traditional solution is to use a native type that exposes a Python interface
and has a constant translation cost. NumPy's \texttt{ndarray} is a typical
implementation of this concept and is the keystone of the Scipy and NumPy
packages. Basically, a NumPy \texttt{ndarray} is a raw C pointer that is exposed
at the Python level. As a tool for scientific computing, Pythran supports such a
structure, implemented as a class wrapping a raw pointer.

@@

The \texttt{reduction(\emph{operator}: \emph{list})} directive is used to
characterize some data dependencies when performing a parallel reduction. The
list of supported operators depend on the input and backend languages: Python
has \texttt{min}/\texttt{max} operators but C/C++ does not. % C++11 has...
C/C++ have
\texttt{\&\&} or \texttt{||} while Python does not.\footnote{Although similar,
the \texttt{and} and \texttt{or} keywords have a special meaning in Python.} The
latest OpenMP specifications~\cite{openmp4} describe a way to declare
user-defined reduction but is not implemented in any compiler yet.

@@

The core directive to handle data parallelism is the \texttt{for} directive that
distributes the iteration space of the associated loop among the existing
threads. To be compatible with OpenMP, the loop iteration space has to be
described by a random access iterator with a total order. Integers used as loop
indices in Fortran and C satisfy this conditions, as well as C++ iterators with
the \texttt{random\_access\_tag} trait. But a Python iterator only advances by a
step of one until it is exhausted, in which case it raises an exception: it
behaves as a forward iterator. To be compatible with OpenMP, these iterators
has to be turned into random access iterators.

@@

Generator expressions are generators whose content is built from an other
iterator. For instance \texttt{(x*x for x in l)} successively yields the square
of each element in \texttt{l}. They behave like adaptors: they apply a
particular expression on each value of the iterator. It is also the case of the
\texttt{enumerate} builtin, that yields each element of the enumerated iterator
associated with its index. These generators are random access iterators only if
their input iterator is a random iterator itself.

@@

\begin{lstlisting}[float,label={lst:sample-api-usage},caption={Example of
        OpenMP API usage from Python.}]
import omp
def omp_get_num_threads():
  nthreads, nthreads_lib = 0, -1
  #omp parallel
  if 1
    #omp critical
    nthreads += 1
    #omp single
    nthreads_lib = omp.get_num_threads()
  return nthreads == nthreads_lib
\end{lstlisting}

@@

>>>>

Since its birth in December 1989, the Python language~\cite{rossum97} has proved
to be useful in various domains, ranging from system administration to web
services, thanks to its dynamicity, expressiveness, rich ecosystem and
``batteries included'' standard library. It is also widely used in scientific
computing~\cite{Oliphant2007}, thanks to the \texttt{numpy} module and the
SciPy~\cite{scipy} project which provide a set of scientific and numeric tools,
ranging from linear algebra to ordinary differential equation solvers or generic
algorithms.

@@

The hybrid approach, where an application contains a mix of interpreted and
native code, is getting widespread in the Python landscape.
It is also the recommended approach to write applications that make use of fine
grain parallelism, as Python parallelism support is only suitable for
coarse-grain parallelism.
To allow the user to write native functions without having to write the
boilerplate glue code that turns Python object into native structures back and
forth, the SciPy package provides the \texttt{weave} module. It makes it
possible to bundle C code snippets into Python code. They are then compiled and
loaded at runtime. Other project like PyCuda or PyOpenCL offers the same
convenience to target hardware accelerators.

@@

\begin{lstlisting}[float, label={lst:buffon-python}, caption={Implementation of
sequential and parallel version of the Buffon algorithm in Python.}]
def buffon(darts, _ = 0):
  hits = 0
  for i in xrange (0, darts):
    x, y = random(), random()
    dist = sqrt(pow(x, 2) + pow(y, 2))
    if dist <= 1.0:
      hits += 1.0
  # hits / throws = 1/4 Pi
  pi = 4 * (hits / darts)
  return pi

from threading import Thread
from multiprocessing import Queue
def threaded_buffon(d):
  def work(darts, queue):
    queue.put(buffon(darts))
  n = 4
  q = Queue()
  threads = [
    Thread(target=work, args=(d//n, q)),
    Thread(target=work, args=(d//n, q)),
    Thread(target=work, args=(d//n, q)),
    Thread(target=work, args=(d//n, q)),
  ]
  map(Thread.start, threads)
  map(Thread.join, threads)
  return sum(q.get() for _ in threads)/n

import functools
from multiprocessing import Pool
def multi_buffon(darts):
  n = 4
  p = Pool(n)
  return sum(p.map(functools.partial(buffon,darts/n), range(n)))/n
\end{lstlisting}

Another widely used solution is the module IPython.parallel, which supports many
different styles of parallelism including: single program, multiple data (SPMD)
parallelism, multiple program, multiple data (MPMD) parallelism. message passing
using MPI, task farming, and data parallel.
FIXME: to be continued...

There have also been several approaches to replace the GIL by Transactional
Memories~\cite{Riley2006,Tabba2010} but none of them made its way to the
mainstream interpreters. As a consequence, Python developers need to write
multi-threaded native modules in order to fully benefit from multiple cores.
This leads to a kind of computations referred as \emph{hybrid computations}.

@@

Cython~\cite{cython2010} is an hybrid Python/C dialect. It extends the Python
syntax with typing information, calls to native functions from third party
libraries, and a limited set of parallelism constructs, such as the possibility
to define parallel loops, but no task parallelism. When possible, it unboxes
Python variables to improve performance. Without type annotations, the
performance improvement is not terrific, but given enough type annotation,
Cython can generates code that runs as fast as its C equivalent, while
maintaining a syntax close to Python.

PLW~\cite{dongara2007} and \texttt{Scipy.weave} both propose another approach
that directly mixes Python with C, using raw strings to hold the C code. PLW
also supports parallel directives that are limited to parallel for loops.
PyCUDA and PyOpenCL~\cite{klockner2012} also target accelerators by mixing
Python with kernels embedded as raw strings containing accelerator code.

The main drawback of these approaches % Reviewer 4 not happy with this
is that they require to modify in a
backward-incompatible way the original code. They require to learn a new
dialect, and the long-term preservation of this investment is not ensured.
Moreover, some of these approaches have no fallback if the code were to be
deployed in an environment where the parallelizing tool/module is not available.
For instance, once a code has been ported to PyCUDA and transformed into CUDA
code embedded in Python strings, there is no way back and the developer has to
maintain two versions of the algorithm.

@@

Most backward-compatible approaches also % Reviewer 4 not happy with this
require to modify the input program.
They do not extend the Python language, but restrict it. As a consequence, they
remain compatible with the original language and do not suffer from the
drawbacks of the previous approaches. They also benefit from existing tools
associated to the language.

@@

At the AST level, it means that the Python AST is first reduced to a tree where
all nodes not available in C++ are transformed into a compatible representation.
For instance, list comprehension expressions are turned into function calls or
tuple unpacking is turned in multiple assignments. During this transformation
process, an expression is always transformed into a single expression and a
statement is always transformed into a single statement.
% FIXME as noted by reviewer 4, the previous sentence is not exact. Just a few
% lines above we said that "tuple unpacking is turned in multiple assignments"
Then this AST is converted into a C++ AST meant to be pretty-printed.

@@

The traditional solution is to use a native type that exposes a Python interface
and has a constant translation cost. NumPy's \texttt{ndarray} is a typical
implementation of this concept and is the keystone of the Scipy and NumPy
packages. Basically, a NumPy \texttt{ndarray} is a raw C pointer that is exposed
at the Python level. The PEP 3118 defined the API that allows to share
efficiently the embeded data between Python and the native world without
involving any copy. As a tool for scientific computing, Pythran supports such a
structure, implemented as a class wrapping a raw pointer.

@@

The \texttt{reduction(\emph{operator}: \emph{list})} directive is used to
characterize some data dependencies when performing a parallel reduction. The
list of supported operators depend on the input and backend languages: Python
has \texttt{min}/\texttt{max} operators but C/C++ does not. % C++11 has...
C/C++ have
\texttt{\&\&} or \texttt{||} while Python does not.\footnote{Although similar,
the \texttt{and} and \texttt{or} keywords have a special meaning in Python.}
% FIXME reviewer 4 doesn't seem to agree with and/or C++ vs Python
The latest OpenMP specifications~\cite{openmp4} describe a way to declare
user-defined reduction but is not implemented in any compiler yet.

@@

The core directive to handle data parallelism is the \texttt{for} directive that
distributes the iteration space of the associated loop among the existing
threads. To be compatible with OpenMP, the loop iteration space has to be
described by a random access iterator with a total order. Integers used as loop
indices in Fortran and C satisfy this conditions, as well as C++ iterators with
the \texttt{random\_access\_tag} trait. But a Python iterator only advances by a
step of one until it is exhausted,
% reviewer 4 disagree with the previous statement
in which case it raises an exception: it
behaves as a forward iterator. To be compatible with OpenMP, these iterators
has to be turned into random access iterators.

@@

Generator expressions are generators whose content is built from another
iterator. For instance \texttt{(x*x for x in l)} successively yields the square
of each element in \texttt{l}. They behave like adaptors: they apply a
particular expression on each value of the iterator. It is also the case of the
\texttt{enumerate} builtin, that yields each element of the enumerated iterator
associated with its index. These generators are random access iterators only if
their input iterator is a random iterator itself.

@@

\begin{lstlisting}[float,label={lst:sample-api-usage},caption={Example of
        OpenMP API usage from Python.}]
import omp
def omp_get_num_threads():
  nthreads, nthreads_lib = 0, -1
  #omp parallel
  if 1:
    #omp critical
    nthreads += 1
    #omp single
    nthreads_lib = omp.get_num_threads()
  return nthreads == nthreads_lib
\end{lstlisting}

@@

% FIXME reviewer 4
% Fig 2, 3, etc: Please use a different color scheme, or add plot markers to
% the plots for the colorblind among us who can't distinguish thin green and
% red lines from each other.

% FIXME reviewer 4
% Fig 2, 3: I believe a log-log plot or a plot of the overall speedup versus
% number of active threads would be a better way to display the data.

% Remaining reviewer's comments:
% - Is it the intent of the Pythran authors to expose the entirety of the
%   OpenMP API, or just a subset?  Task parallelism?
% - V.C: "Cython is not backwards compatible with Python".  This is false:
%   Cython's Pure Python mode is fully backwards compatible.
% - Figs 5,6,7,8: to do apples-to-apples, what about Cython + prange?
% - Perhaps the only complain is that I would have liked to see the timing for
%   compilation to be included in the article.  Pythran and Cython are very
%   different beasts than Numba or Parakeet in this sense, and the later maybe
%   more useful to compile code that still not exists at compilation time (like
%   user queries, for example).
